Universal algebra and the \texorpdfstring{$ \lambda $}{lambda}-calculus

### 2024-06-17 report/src/algebraic-structures.tex
Show that the pullback is essentially surjective
Show that the algebra pullback is functorial in the

\begin{definition}
  A \textit{morphism}\index{algebraic theory!morphism} $ F $ between algebraic theories $ T $ and $ T^\prime $ is a sequence of functions $ F_n: T_n \to T^\prime_n $ (we usually leave the $ n $ implicit) such that
  \begin{align*}
    F_n(x_j) &= x_j\\
    F_n(f \bullet g) &= F_m(f) \bullet (F_n(g_i))_i
  \end{align*}
  for all $ 1 \leq j \leq n $, $ f: T_m $ and $ g: T_n^m $.
\end{definition}

@@

\begin{definition}
  For an algebraic theory $ T $, a \textit{morphism}\index{algebra!morphism} $ F $ between $ T $-algebras $ A $ and $ A^\prime $ is a function $ F: A \to A $ such that
  \[ F(f \bullet a) = f \bullet (F(a_i))_i \]
  for all $ f: T_n $ and $ a: A^n $.
\end{definition}

@@

Actually, we can recover some information about the algebraic theory morphism from this pullback functor. For example,
\begin{lemma}\label{lem:algebra-pullback-equivalence}
  If $ F^* $ is an equivalence of categories, $ F $ is an isomorphism.
\end{lemma}
\begin{proof}
  \TODO
\end{proof}

@@

\begin{definition}
  For an algebraic theory $ T $, a \textit{morphism}\index{presheaf!morphism} $ F $ between $ T $-presheaves $ P $ and $ P^\prime $ is a sequence of functions $ F_n: P_n \to P^\prime_n $ such that
  \[ F_n(t \bullet f) = F_m(t) \bullet f \]
  for all $ t: P_m $ and $ f: T_n^m $.
\end{definition}

@@

\begin{definition}[$ \lambda $-theory morphism]
  A \textit{morphism}\index{$ \lambda $-theory!morphism} $ F $ between $ \lambda $-theories $ L $ and $ L^\prime $ is an algebraic theory morphism $ F $ such that
  \begin{align*}
    F_n(\lambda_n(f)) &= \lambda_n(F_{n + 1}(f))\\
    \rho_n(F_n(g)) &= F_{n + 1}(\rho_n(g))
  \end{align*}
  for all $ f: L_{n + 1} $ and $ g: L_n $.
\end{definition}

@@

  Lastly, the functor is essentially surjective, since we can lift a coslice $ f: S \to X $ to a $ F(S) $-algebra $ X $, with action
  \[ i \bullet x = x_i \quad \text{and} \quad s \bullet x = f(s). \]

@@

$ F(\emptyset) $ is, in some sense, the smallest nontrivial algebraic theory. Then $ F(S) $ is the smallest nontrivial algebraic theory that has the elements of $ S $ as constants.

@@

\begin{remark}
  Recall from \ref{lem:lawvere-algebra} that $ T $-algebras are equivalent to finite-product-preserving $ \SET $-valued functors on the Lawvere theory $ \mathbf L $ associated to $ T $. Now, recall from (\TODO) that we can embed any category inside its category of $ \SET $-valued functors. This embeds $ n: \mathbf L $ as the theory algebra $ T_n $. Then the Yoneda Lemma gives a bijection
  \[ \Alg T(T_n, A) \cong A^n = \SET(\{ 1, \dots, n \}, A) \]
  natural in $ n $ and $ A $. Explicitly, it sends $ F: \Alg T(T_n, A) $ to $ i \mapsto F(x_i) $ and $ (a_i)_i : A^n $ to $ f \mapsto f \bullet a $. The natural equivalence immediately shows that $ T_n $ is the free $ T $-algebra on $ n $ elements.
\end{remark}

>>>>

\begin{definition}
  A \textit{morphism}\index{algebraic theory!morphism} $ f $ between algebraic theories $ T $ and $ T^\prime $ is a sequence of functions $ f_n: T_n \to T^\prime_n $ (we usually leave the $ n $ implicit) such that
  \begin{align*}
    f_n(x_j) &= x_j\\
    f_n(f \bullet g) &= f_m(f) \bullet (f_n(g_i))_i
  \end{align*}
  for all $ 1 \leq j \leq n $, $ f: T_m $ and $ g: T_n^m $.
\end{definition}

@@

\begin{definition}
  For an algebraic theory $ T $, a \textit{morphism}\index{algebra!morphism} $ f $ between $ T $-algebras $ A $ and $ A^\prime $ is a function $ f: A \to A $ such that
  \[ f(t \bullet a) = f \bullet (f(a_i))_i \]
  for all $ t: T_n $ and $ a: A^n $.
\end{definition}

@@

Actually, we can recover some information about the algebraic theory morphism from this pullback functor. For example,
\begin{lemma}\label{lem:algebra-pullback-equivalence}
  If $ f^* $ is an equivalence of categories, $ f $ is an isomorphism.
\end{lemma}
\begin{proof}
  \TODO
\end{proof}

@@

\begin{definition}
  For an algebraic theory $ T $, a \textit{morphism}\index{presheaf!morphism} $ f $ between $ T $-presheaves $ P $ and $ P^\prime $ is a sequence of functions $ f_n: P_n \to P^\prime_n $ such that
  \[ f_n(t \bullet f) = f_m(t) \bullet f \]
  for all $ t: P_m $ and $ f: T_n^m $.
\end{definition}

@@

\begin{definition}[$ \lambda $-theory morphism]
  A \textit{morphism}\index{$ \lambda $-theory!morphism} $ f $ between $ \lambda $-theories $ L $ and $ L^\prime $ is an algebraic theory morphism $ f $ such that
  \begin{align*}
    f_n(\lambda_n(s)) &= \lambda_n(f_{n + 1}(s))\\
    \rho_n(f_n(t)) &= f_{n + 1}(\rho_n(t))
  \end{align*}
  for all $ s: L_{n + 1} $ and $ t: L_n $.
\end{definition}

@@

  Lastly, the functor is essentially surjective, since we can lift an object $ f: S \to X $ to a $ F(S) $-algebra $ X $, with action
  \[ i \bullet x = x_i \quad \text{and} \quad s \bullet x = f(s). \]

@@

\begin{remark}
  $ F(\emptyset) $ is, in some sense, the smallest nontrivial algebraic theory. Then $ F(S) $ is the smallest nontrivial algebraic theory that has the elements of $ S $ as constants.
\end{remark}

\begin{remark}
  Note that the category of $ F(\emptyset) $-algebras is equivalent to the coslice-category $ (\emptyset \downarrow \SET) $, which, since $ \emptyset $ is the initial set, is just equivalent to $ \SET $.
\end{remark}

@@

\begin{remark}
  Recall from \ref{lem:lawvere-algebra} that $ T $-algebras are equivalent to finite-product-preserving $ \SET $-valued functors on the Lawvere theory $ \mathbf L $ associated to $ T $. Now, recall from (\TODO) that we can embed any category inside its category of $ \SET $-valued functors. This embeds $ n: \mathbf L $ as the theory algebra $ T_n $. Then the Yoneda Lemma gives a bijection
  \[ \Alg T(T_n, A) \cong A^n = \SET(\{ 1, \dots, n \}, A) \]
  natural in $ n $ and $ A $. Explicitly, it sends $ f: \Alg T(T_n, A) $ to $ i \mapsto f(x_i) $ and $ (a_i)_i : A^n $ to $ f \mapsto f \bullet a $. The natural equivalence immediately shows that $ T_n $ is the free $ T $-algebra on $ n $ elements.
\end{remark}

Using some additional machinery, we can combine this with the algebra pullback functor to get another functor:
\begin{definition}\label{def:theory-algebra-pullback-functor}
  Take a nonnegative integer $ n $. Take an object $ S : \AlgTh $. For every object $ T : \AlgTh $, we can take the $ T $-algebra $ T_n $. Then every morphism $ f: \AlgTh(S, T) $ gives an $ S $-algebra $ f^* T_n $. In fact, this is a functor from $ (S \downarrow \AlgTh) $ to $ \Alg S $: We send a morphism $ g: (S \downarrow \AlgTh)((T, f), (T^\prime, f^\prime)) $ to
  \[ g_n: f^* T_n \to (f^\prime)^* T^\prime_n. \]
  This is an algebra morphism because it commutes with $ f $, $ f^\prime $ and the substitution of $ T $ and $ T^\prime $: For $ s: S_m $ and $ t: (f^* T_n)^m $, we have
  \[
      g_n(s \bullet_{f^* T_n} t)
    = g_m(f_m(s)) \bullet_{T^\prime} (g_n(t_i))_i
    = s \bullet_{(f^\prime)^* T^\prime_n} (g_n(t_i))_i.
  \]
  The functor obviously preserves the identity morphisms and composition of morphisms, so it is indeed a functor.
\end{definition}

### 2024-06-17 report/src/paper.tex
Show that the algebra pullback is functorial in the

\section{An elementary proof of the `Fundamental Theorem'}\label{sec:elementary-fundamental-theorem}
The final theorem that Hyland works towards in his paper constructs an equivalence between $ \lambda $-calculi and $ \Lambda $-algebras. In this section, we will provide an elementary proof of this theorem.

@@

In both cases, one part of the equivalence is very easy:
\begin{definition}
  For all $ n $, we have a functor from lambda theories to $ \Lambda $-algebras. It sends the $ \lambda $-theory $ L $ to the $ L $-algebra $ L_n $ and then turns this into a $ \Lambda $-algebra via the morpism $ \Lambda \to L $. It sends morphisms $ f: L \to L^\prime $ to the algebra morphism $ f_n : L_n \to L^\prime_n $.
\end{definition}
Here we are only interested in the case $ n = 0 $.

We will use Lemma \ref{lem:univalen-category-equivalence}, by showing that $ L \mapsto L_0 $ is a weak equivalence.

@@

\TODO show functoriality of pullback in coslice category.

@@

  Now, to show that it is full, take $ B, C: \Alg{E_{\R_A}(U)} $. Also, take a morphism
  \[ h : \Alg{\Lambda_A}([a \mapsto \epsilon_A^{-1}(a) \bullet_B (), \iota_{\Lambda_n}]_n^* B, [a \mapsto \epsilon_A^{-1}(a) \bullet_C (), \iota_{\Lambda_n}]_n^* C). \]
  By Remark \ref{rem:endomorphism-theory-algebra-to-endomorphism-theory-algebra}, and by functoriality of the endomorphism theory construction, we get a commutative diagram
  \begin{center}
    \begin{tikzcd}
      & E_{\R_A}(U) \arrow[ld, "a \mapsto \epsilon_A^{-1}(a) \bullet_B ()"'] \arrow[rd, "a \mapsto \epsilon_A^{-1}(a) \bullet_C ()"]\\
      E_{\R_{\iota_{\Lambda}^* B}}(U) \arrow[rr, "(h)_n"] & & E_{\R_{\iota_{\Lambda}^* C}}(U)
    \end{tikzcd}
  \end{center}
  Now, using functoriality of the pullback of the theory algebras in the coslice category $ (E_{\R_A}(U) \downarrow \AlgTh) $, and using the previous lemma, we get the following diagram of $ E_{\R_A}(U) $-algebras:
  \begin{center}
    \begin{tikzcd}[sep = large]
      (a \mapsto \epsilon_A^{-1}(a) \bullet_B ())^* (E_{\R_{\iota_\Lambda^* B}}(U)_0)
        \arrow[r, "(a \mapsto \epsilon_A^{-1}(a) \bullet ())^* h"]
        \arrow[d, "\epsilon_{\iota_\Lambda^* B}", "\sim"'] &
      (a \mapsto \epsilon_A^{-1}(a) \bullet_C ())^* (E_{\R_{\iota_\Lambda^* C}}(U)_0)
        \arrow[d, "\epsilon_{\iota_\Lambda^* C}", "\sim"']\\
      B \arrow[r, dashed, "\bar h"] & C
    \end{tikzcd}
  \end{center}
  This gives us the lift
  \[ \bar h = \epsilon_{\iota_\Lambda^* B}^{-1} \cdot (a \mapsto \epsilon_A^{-1}(a) \bullet ())^* h \cdot \epsilon_{\iota_\Lambda^* C} : \Alg{E_{\R_A}(U)}(B, C) \]

>>>>

\section{An elementary proof of the `Fundamental Theorem'}\label{sec:elementary-fundamental-theorem}
The final theorem that Hyland works towards in his paper constructs an equivalence between $ \lambda $-calculi and $ \Lambda $-algebras. In this section, we will provide an elementary proof of this theorem. Then, in the next section, we will give Hyland's original proof, which uses more high-level categorical reasoning.

@@

In both cases, one part of the equivalence is very easy. Recall that the pure $ \lambda $-calculus, the $ \lambda $-theory $ \Lambda $, is the initial object of $ \LamTh $. That means that $ \LamTh $ is equivalent to $ (\Lambda \downarrow \LamTh) $. Now, using Definition \ref{def:theory-algebra-pullback-functor}, for any $ n $ we get a functor
\[ -_n : \LamTh \to \Alg \Lambda. \]
Here we are only interested in the case $ n = 0 $.

We will use Lemma \ref{lem:univalen-category-equivalence}, by showing that $ -_0 $ is a weak equivalence.

@@

@@

  Now, to show that it is full, take $ B, C: \Alg{E_{\R_A}(U)} $. Also, take a morphism
  \[ h : \Alg{\Lambda_A}([a \mapsto \epsilon_A^{-1}(a) \bullet_B (), \iota_{\Lambda_n}]_n^* B, [a \mapsto \epsilon_A^{-1}(a) \bullet_C (), \iota_{\Lambda_n}]_n^* C). \]
  By Remark \ref{rem:endomorphism-theory-algebra-to-endomorphism-theory-algebra}, and by functoriality of the endomorphism theory construction, we get a commutative diagram
  \begin{center}
    \begin{tikzcd}
      & E_{\R_A}(U) \arrow[ld, "a \mapsto \epsilon_A^{-1}(a) \bullet_B ()"'] \arrow[rd, "a \mapsto \epsilon_A^{-1}(a) \bullet_C ()"]\\
      E_{\R_{\iota_{\Lambda}^* B}}(U) \arrow[rr, "(h)_n"] & & E_{\R_{\iota_{\Lambda}^* C}}(U)
    \end{tikzcd}
  \end{center}
  Now, using Definition \ref{def:theory-algebra-pullback-functor} for the coslice category $ (E_{\R_A}(U) \downarrow \AlgTh) $, and using the previous lemma, we get the following diagram of $ E_{\R_A}(U) $-algebras:
  \begin{center}
    \begin{tikzcd}[sep = large]
      (a \mapsto \epsilon_A^{-1}(a) \bullet_B ())^* (E_{\R_{\iota_\Lambda^* B}}(U)_0)
        \arrow[r, "(a \mapsto \epsilon_A^{-1}(a) \bullet ())^* h"]
        \arrow[d, "\epsilon_{\iota_\Lambda^* B}", "\sim"'] &
      (a \mapsto \epsilon_A^{-1}(a) \bullet_C ())^* (E_{\R_{\iota_\Lambda^* C}}(U)_0)
        \arrow[d, "\epsilon_{\iota_\Lambda^* C}", "\sim"']\\
      B \arrow[r, dashed, "\bar h"] & C
    \end{tikzcd}
  \end{center}
  This gives us the lift
  \[ \bar h = \epsilon_{\iota_\Lambda^* B}^{-1} \cdot (a \mapsto \epsilon_A^{-1}(a) \bullet ())^* h \cdot \epsilon_{\iota_\Lambda^* C} : \Alg{E_{\R_A}(U)}(B, C) \]
